---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the research question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?
* How many bookings were canceled?
* What was the booking ratio between resort hotels and city hotels?
* What was the percentage of booking for each year?
* Which was the busiest month for hotels?
* Most guests came from which country?
* How long do most people stay in hotels?
* Which was the most booked accommodation type (Single, Couple, Family)?

### c) Understanding the context
You are a Data Scientist working for Hilton International Hotels and need to help the management decide on
the strategies that will lead to revenue growth.
You have a dataset containing information on the various chain of hotels, including when customers made
the booking, length of stay, the number of adults, children, or babies, and the number of available parking
spaces, among other things.

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.
*Loading the relevant libraries
*Reading the data
*Exploring the data to understand it better
*External Data Source Validation
*cleaning the data
*Sorting the data
*Analyzing the data to come up with inferences
*Visualizing the data
*Provide recommendations
### e) Data Relevance
How relevant was the provided data?
It was sufficient for the Analysis

## 1. Loading the relevant libraries
```{r}
library(tidyverse)
```

## 2. Reading and exploring the data
```{r}
# Load the data below
# --- 
# Dataset url = https://bit.ly/2WvQbhJ
# --- 
# YOUR CODE GOES BELOW
Bookings_df<- read.csv("hotel_bookings.csv")
Bookings_df
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(Bookings_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(Bookings_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(Bookings_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#
dim(Bookings_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(Bookings_df)
```
Record your general observations below:

The data set has 119390 rows and 32 columns
Some columns have null values 



## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
str(Bookings_df)
```
```{r}
colSums(is.na(Bookings_df))
```

We observe the following from our dataset:

Children variable has 4 missing values
Observation 2

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
Bookings_df[duplicated(Bookings_df), ]
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
#if (any(is.na(Bookings_df[,hotel]))) {next}
colSums(is.na(Bookings_df))
```

We observe the following from our dataset:

Children has 4 missing values

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
#
is.na(Bookings_df)
```

We observe the following from our dataset:

There are no rows with all null

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 
install.packages("reshape2")
library(reshape2)
# Select the desired columns
Bookings_df <- Bookings_df %>%
na.omit() %>%
  select()
Bookings_df
# Create a correlation matrix
corr_matrix <- cor(Bookings_df, method="s")
head(corr_matrix)
# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df
# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

We observe the following from our dataset:
No correlation 

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
Bookings_df1 = select(Bookings_df, -c(company))
names(Bookings_df1)
```

From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the missing values i.e. for TO
# ---
# YOUR GOES BELOW
#
Bookings_df$distribution_channel[is.na(Bookings_df$distribution_channel)] <- "TO"
Bookings_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#
adults <-Bookings_df1$adults
babies <-Bookings_df1$babies
child <-Bookings_df1$children
Bookings_df2 <- subset(Bookings_df, adults!="0" & babies!="0" & child!="0")
head(Bookings_df2)
#Bookings_df2 <- Bookings_df1[-c(adults == 0), ]
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
Bookings_df2$children[is.na(Bookings_df2$children)] <- mean(dt$age, na.rm = TRUE)
Bookings_df2
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
is.na(Bookings_df2)
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
lapply(Bookings_df2$children,as.numeric)
lapply(Bookings_df2$company,as.numeric)
lapply(Bookings_df2$agent,as.numeric)
#  Bookings_df <- Bookings_df %>% 
#     mutate_at(c(11, 24, 25), as.integer())
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot
# Create the summary table
cancelled_df <- hotel_df %>%
  group_by(hotel) %>%
  summarise(is_canceled = sum(is_canceled))
# Plotting Barplot chart to visualize the data
ggplot(cancelled_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Cancelled Bookings per Hotel", 
    x = "hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = is_canceled, face = "bold"), vjust = 1.5, colour = "black")
```
City hotel had 33102 cancelled bookings while resort hotel had 11122 cancelled bookings

```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel
# Create the summary table
ratio_df <- Bookings_df %>%
  group_by(hotel) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(ratio_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Hotel", 
    x = "hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

City hotel had a ratio of 61.5% and resort hotel had a ratio of 38.5%
```{r}
# 3. What was the percentage of booking for each year?
# ---
#
year_df <- Bookings_df %>%
  group_by(arrival_date_year) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(year_df, aes(x = arrival_date_year, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Year", 
    x = "year", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
2016 had the highest bookings at 48.4% followed by 2017 at 33.2% and finally 2015 at 18.4% 
```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

Most bookings are in the months of July and August
```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#
countries_df <- Bookings_df %>%
  group_by(country) %>%
  summarise(is_canceled = sum(is_canceled==0))
countries_df
arrange(countries_df,desc(is_canceled))
#Filtering to get the top three countries
top3_countries_df <- filter(countries_df, country == "PRT" | country == "GBR" | country == "FRA")
top3_countries_df
# Plotting Barplot chart to visualize the data
ggplot(top3_countries_df, aes(x = country, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "country", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
PRT had the highest number of customers booking, followed by GBR and FRA in the third position
```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
#
stays_df <- Bookings_df %>%
  mutate(total_stays = stays_in_weekend_nights + stays_in_week_nights) %>%
  group_by(total_stays) %>%
  summarise(is_canceled = sum(is_canceled==0))
  arrange(stays_df, desc(is_canceled))
stays_df
# Plotting Barplot chart to visualize the data
ggplot(stays_df, aes(x = total_stays)) +
  geom_histogram(
    bins = 10,fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Total stays in nights", 
    x = "total_stays", 
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  scale_x_continuous(breaks = seq(0, 50,by=5))
```
Most people stay between 0 to 3 days

```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation_df <- Bookings_df %>%
  mutate(hotel_accommodation =
           case_when(
             Bookings_df$adults >= 0 & (Bookings_df$children >= 1 | Bookings_df$babies >= 1) ~ "Family",
             Bookings_df$adults == 2 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Couple",
             Bookings_df$adults == 1 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Single",
            Bookings_df$adults > 2 & (Bookings_df$children == 0 | Bookings_df$babies == 0) ~ "Family")
  )%>%
  filter(is_canceled == 0)
  
accommodation_df1 <- accommodation_df %>%
  group_by(hotel_accommodation) %>%
  count(hotel_accommodation)
#Plotting the bar chart
ggplot(accommodation_df1, aes(x = hotel_accommodation, y = n, fill = accommodation_df)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Hotel Accommodataion by accomodation type",
    x = "Hotel Accommodation", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Couples make up the largest bookings, followed by singles and Family

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
In the months of January, December and November to try and boost bookings


### 5.b) Recommendations
From the above analysis, below are our recommendations:
More offers should be considered in the months of January, December and November to try and boost bookings
More family products should be tailored so as to encourage more bookings from that segment
Offers and discounts around long stays should be considered as longer stays increase revenue
Capacity building should be done for the months of July and August to cater for the many bookings

## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data? Yes we had the right data
b). Do we need other data to answer our question? No
c). Did we have the right question? Yes we had the right questions

```{r}
library(htmltools)
```